---
title: "Lithuanian NVI Biobank Stored Samples"
author: "Simona Zubavičiūtė"
date: "2024-02-29"
output: 
  html_document:
    df_print: paged
---

# Preparation for data analysis and data cleaning

## Importing libraries

```{r, message = FALSE, warning = FALSE}
library(dplyr)
library(tidyverse)
library(base)
library(lubridate)
library(ggplot2)
library(webr)
library(plotly)
```

## Importing data

```{r, R.options=list(max.print=100)}
data <- read.table(file = "C:\\Users\\Simona\\Desktop\\Personal_Portfolio\\Biobank_LT\\biobank.txt", header = TRUE, sep = ',')

data
```

## Checking for the number of rows and columns in data set

```{r}
glimpse(data)
```
## Merging the prefix and suffix to form ID 

```{r, R.options=list(max.print=100)}
data$ID <- paste(data$emimprefix, data$megnrsuffix, sep = "")

data
```

## Cheking if all samples have unique ID

```{r, R.options=list(max.print=100)}
unique(data$ID)
```

## Removing duplicates to get clean data set

```{r, R.options=list(max.print=100)}
df <- data %>% distinct()

df
```

## Renaming columns from Lithuanian to English

```{r, R.options=list(max.print=100)}
df_renamed <-
df %>% 
  rename(age_group = amzgr,
    sex = lytis,
    diagnosis = dgn,
    localization = lokalizacija,
    stored_sample = megpavad,
    container = indpavad,
    amount = kiekis,
    measurmement = vnt,
    year = megmetai
  )

df_renamed
```
## Changing localization name from Lithuanian to English

```{r, R.options=list(max.print=100)}

df_renamed_rv <-
df_renamed %>%
  mutate(localization = recode(localization,
                            "Gimdos kūnas" = "Uterus", 
                            "Inkstai" = "Kidneys",
                            "Kiaušidės" = "Ovaries", 
                            "Krūtis" = "Breast", 
                            "Melanoma" = "Melanoma", 
                            "Plaučiai" = "Lungs", 
                            "Prostatos biopsija" = "Prostate", 
                            "Sėklidės" ="Testicles", 
                            "Žarnynas" ="Intestine"))
df_renamed_rv

```

## Removing not useful columns 

```{r, R.options=list(max.print=100)}
df_full <- subset(df_renamed_rv, select = -c(megmen, emimprefix, megnrsuffix))

df_full
```
## Changing diagnosis values to Cancerous and Non-Cancerous

```{r, R.options=list(max.print=100)}
df_full$diagnosis <- sub(".\\d+","Cancerous", df_full$diagnosis)

df_full
```
```{r, R.options=list(max.print=100)}
df_full$diagnosis <- sub("^$","Non-cancerous", df_full$diagnosis)

df_full
```

# Visualization

## Visualizing what are the most common localizations of stored samples by age group

```{r}
breaks_y = c(0, 500, 1000, 1500, 2000, 2500, 3000, 3500, 4000)

plot1 <- ggplot(df_full, aes(x = age_group, fill = localization)) +
  geom_bar() +
  labs(x = "Age group", y = "Counts per Localization", title = "Stored Samples by Age Group at NVI Biobank", color = "Localization", fill = "Localization") +
  scale_y_continuous(limits = c(0, 4000), breaks = breaks_y) +
  theme_bw() +
  theme(axis.text = element_text(size = 9),
    axis.title.y = element_text(size = 12, face ="bold"),
    axis.title.x = element_text(size = 12, face ="bold"),
    legend.title=element_text(size=10, face ="bold"),
    legend.text=element_text(size=10),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 12, face = "bold"))

ggplotly(plot1)

```


## Visualizing how stored sample count changed during the years 2018-2022 by sex and localization

```{r}
breaks_y = c(0, 1000, 2000, 3000, 4000)

df_full_mod <- df_full %>%
  mutate(sex = recode(sex, "V" = "Male", "M" = "Female"))

plot2 <- ggplot(df_full_mod, aes(x = year, fill = localization)) +
  geom_bar() + 
  facet_wrap(vars(sex)) +
  labs(x = "Year of Collection", y = "Counts per Localization", title = "Stored Samples by Sex at NVI Biobank", color = "Localization", fill = "Localization") +
  scale_y_continuous(limits = c(0, 4000, breaks = breaks_y)) +
  scale_fill_discrete(labels=c("Gimdos kūnas" = "Uterus", 
     "Inkstai" = "Kidneys",
     "Kiaušidės" = "Ovaries", 
     "Krūtis" = "Breast", 
     "Melanoma" = "Melanoma", 
     "Plaučiai" = "Lungs", 
     "Prostatos biopsija" = "Prostate biopsy", 
     "Sėklidės" ="Testicles", 
     "Žarnynas" ="Intestine")) +
  theme_bw() +
  theme(axis.text = element_text(size = 9),
    axis.title.y = element_text(size = 12, face ="bold"),
    axis.title.x = element_text(size = 12, face ="bold"),
    legend.title=element_text(size=10, face ="bold"),
    legend.text=element_text(size=10),
    plot.title = element_text(size = 12, face = "bold")) 
```

```{r, include = FALSE}
plot2_flip <- plot2 + coord_flip()
```

```{r}
ggplotly(plot2_flip) 
```



## Visualizing whether stored samples by localization and age group were cancerous or not

```{r}
breaks_y = c(0, 500, 1000, 1500, 2000, 2500, 3000, 3500, 4000)

plot3 <- ggplot(df_full, aes(x = age_group, fill = localization)) +
  geom_bar()  +
  facet_wrap(vars(diagnosis)) +
  labs(x = "Age group", y = "Counts per Localization", title = "Stored Samples by Diagnosis and Age Group at NVI Biobank", color = "Localization", fill = "Localization") +
  scale_y_continuous(limits = c(0, 4000), breaks = breaks_y) +
  scale_fill_discrete(labels=c("Gimdos kūnas" = "Uterus", 
     "Inkstai" = "Kidneys",
     "Kiaušidės" = "Ovaries", 
     "Krūtis" = "Breast", 
     "Melanoma" = "Melanoma", 
     "Plaučiai" = "Lungs", 
     "Prostatos biopsija" = "Prostate biopsy", 
     "Sėklidės" ="Testicles", 
     "Žarnynas" ="Intestine")) +
  theme_bw() +
  theme(axis.text = element_text(size = 9),
    axis.title.y = element_text(size = 12, face ="bold"),
    axis.title.x = element_text(size = 12, face ="bold"),
    legend.title=element_text(size=10, face ="bold"),
    legend.text=element_text(size=10),
    axis.text.x = element_text(angle = 90, hjust = 1),
    plot.title = element_text(size = 12, face = "bold"))

ggplotly(plot3) 
```
